---
title: "Designing and Creating a Database"
Author: Mahmood Hossain Farsim 
output: html_notebook
---

In this guided project, we're going to learn how to:

    Import data into SQLite
    Design a normalized database schema
    Create tables for our schema
    Insert data into our schema


We will work with data from Major League Baseball games compiled by Retrosheet, a non-profit organization that's gathered game statistics going back to the 1800s to today. The main file we will work from is game_log.csv

In addition to the main file, we also have three "helper" files sourced from Retrosheet:

    park_codes.csv
    person_codes.csv
    team_codes.csv


## STEP 1 -Exploring The Data 


```{r}
library(tidyverse)
library(RSQLite)
library(DBI)


log <- read_csv("game_log.csv", 
                col_types = cols(.default = "c",
                                 v_league = "c", h_league = "c",
                                 `3b_umpire_id` = "c", `3b_umpire_name` = "c",
                                 `2b_umpire_id` = "c", `2b_umpire_name` = "c",
                                 `lf_umpire_id` = "c", `lf_umpire_name` = "c",
                                 `rf_umpire_id` = "c", `rf_umpire_name` = "c",
                                 completion = "c", winning_rbi_batter_id = "c",
                                 winning_rbi_batter_id_name = "c", protest = "c",
                                 v_first_catcher_interference = "c", 
                                 h_first_catcher_interference = "c"))
dim(log)
head(log)
```

We can see that our log file consists of more than 170000 game entries. There are 161 data points (columns) in the log. We need to look at this very closely and keep on researching the attributes and connections and everything around this information. Now lets read other helper files.

```{r}
park <- read.csv("park_codes.csv")
dim(park)
head(park)
``` 

The Park Codes csv file inclueds the venue information. Probably connected to the park_id field in the main log file.


```{r}
person <- read.csv("person_codes.csv")
dim(person)
head(person)
```

So here we have some names. may be these people plays the game, then at some point they debut as managers and coaches. Probably connected to the main log file via the umpire name or player names etc.




```{r}
team <- read.csv("team_codes.csv")
dim(team)
head(team)
```
SO, here we have information of a total of 150 teams. Maybe we will find these team names in the main game log file. Maybe connected to the main log file through the 'league' field through the park file which also have a league column. EXPLORE THE HELL OF IT ! MORE >>ENDLESS>>>>MICRO DETAILS>>>MORE MORE MORE>>>>NEW THOUGHTS



**What each defensive position number represents in the main log file** : 

There are total of 9 defensive positions. Each position number mean a particular location  where a player stands on the field. In the main log file, we can see there are home(h) and visitor (v) players id, name and defensive positions are mentioned. 

These are the defensive positiones :


1. Pitcher
2. Catcher
3. First Baseman
4. Second Baseman
5. Third baseman
6. Shortstop
7. Left Fielder
8. Center Fielder
9. Right Fielder

**The values in the various league fields, and which leagues they represent** : We see the league fields are in "Team" and "Park" data frames.Also, there are home league and visitor league fields in the main log file. 


AL : American League
NL : National League 
UA : Union Associations
PL : Players League
FL : Federal League
AA : American Associations 

## STEP 2 -Importing Data into SQlite


The RSQLite library itself gives us a few ways to bring the data into a SQLite database. By far, the easiest way is to load the data into a variable, use the dbWriteTable() method and pass the variable into the value argument. 

```{r}
conn <- dbConnect(SQLite(), "mlb.db")
dbWriteTable(conn = conn, name = "game_log", 
             value = log, row.names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "person_codes", 
             value = person, row.names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "team_codes", 
             value = team, row.names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "park_codes", 
             value = park, row.names = FALSE, header = TRUE)


```

```{r}
# Confirm that all of the tables are in 
dbListTables(conn)
```

Create a new column in the game_log table called game_id:

    
    Once you write the query, you can use dbExecute() to perform the query
    
    
    Now we will add a column called "game_id" to the main log file


```{r}
alter_game_log_command <- "
  ALTER TABLE game_log
  ADD COLUMN game_id TEXT;
"

dbExecute(conn, alter_game_log_command)
```

Use SQL string concatenation to update the new columns with a unique ID using the Retrosheet format outlined above.

```{r}
update_game_log_command <- "
  UPDATE game_log
  SET game_id = date || h_name || number_of_game
  ;
  "

dbExecute(conn, update_game_log_command)
```

```{r}
# Make sure that our queries worked
check_query_1 <- "
  SELECT
      game_id,
      date,
      h_name,
      number_of_game
  FROM game_log
  LIMIT 5;
"
check <- dbGetQuery(conn, check_query_1)
head(check)
```



# Looking for Normalization Opportunities

The following are opportunities for normalization of our data:

* In person_codes, all the debut dates will be able to be reproduced using game log data.
* In team_codes, the start, end and sequence columns will be able to be reproduced using game log data.
* In park_codes, the start and end years will be able to be reproduced using game log data. While technically the state is an attribute of the city, we might not want to have a an incomplete city/state table so we will leave this in.
* There are lots of places in game log where we have a player ID followed by the players name. We will be able to remove this and use the name data in person_codes
* In game_log, all offensive and defensive stats are repeated for the home team and the visiting team. We could break these out and have a table that lists each game twice, one for each team, and cut out this column repetition.
* Similarly, in game_log, we have a listing for 9 players on each team with their positions - we can remove these and have one table that tracks player appearances and their positions.
* We can do a similar thing with the umpires from game_log, instead of listing all four positions as columns, we can put the umpires either in their own table or make one table for players, umpires and managers.
* We have several awards in game_log like winning pitcher and losing pitcher. We can either break these out into their own table, have a table for awards, or combine the awards in with general appearances like the players and umpires.


**Lets take a look at our game_log database**

```{r}
dbGetQuery(conn,"SELECT * FROM game_log; ")
```

**Our goal here is to have a normalized database. Normalized means we wont accept duplications. If needed, we will create new tables that will have the data organized. To normalize the database, go through each table and catagorize them**

Alright, we were asked to design our schema using dbdesigner.net. We were creating the schema but it is consuming a lot of time . However, this is the process in real life. 

TO save time and move forward, we are stepping into the next task where DQ might show their suggested schema. 

![alt text](C:\Users\farsim\Documents\Dataquest Guided Project - Designing and Creating Database\mlb_schema.svg)

Now we will make some new tables as part of the normalization process. We are actually creating a "The more organized database" using and manipulating our resources.

## Create Person, park, league and appearance_type tables 

```{r}
create_person_command <- "
  CREATE TABLE IF NOT EXISTS PERSON (
    person_id TEXT PRIMARY KEY,
    first_name TEXT,
    last_name TEXT
  );
"
#making the table
dbExecute(conn, create_person_command)

#inserting values from person_codes file
insert_to_person <- "
  INSERT OR IGNORE INTO person
  SELECT
      id,
      first,
      last
  FROM person_codes;
"

dbExecute(conn, insert_to_person)

#lets check the table 

check_query_2 <- "SELECT * FROM person LIMIT 5;"
check <- dbGetQuery(conn, check_query_2)
head(check)

```

YESS !! Success. Lets do this for rest of the tables . This is our real goal in this project. 

the dbexecute command creates or modifies a table and dbgetquery performs a query from a table. 

Park Table

```{r}
create_park_command <- "
  CREATE TABLE IF NOT EXISTS park (
      park_id TEXT PRIMARY KEY,
      name TEXT,
      nickname TEXT,
      city TEXT,
      state TEXT,
      notes TEXT
  );
"
dbExecute(conn, create_park_command)
insert_to_park <- "
  INSERT OR IGNORE INTO park
  SELECT
      park_id,
      name,
      aka,
      city,
      state,
      notes
  FROM park_codes;
"
dbExecute(conn, insert_to_park)
check_query_3 <- "SELECT * FROM park LIMIT 5;"
check_3 <- dbGetQuery(conn, check_query_3)

head(check_3)
```

League Table 

```{r}
create_league_command <- "
  CREATE TABLE IF NOT EXISTS league (
      league_id TEXT PRIMARY KEY,
      name TEXT
  );
"
dbExecute(conn, create_league_command)
insert_to_league <- '
  INSERT OR IGNORE INTO league
  VALUES
      ("NL", "National League"),
      ("AL", "American League"),
      ("AA", "American Association"),
      ("FL", "Federal League"),
      ("PL", "Players League"),
      ("UA", "Union Association")
  ;
'
dbExecute(conn, insert_to_league)
check_query_4 <- "SELECT * FROM league"
check <- dbGetQuery(conn, check_query_4)
head(check)
```

Appearance_type table : DQ provided a appearance type.csv file and we gotta make a table from that file. Our appearance table will include data on players with positions, umpires, managers, and awards (like winning pitcher). This table will store information on what different types of appearances are available

```{r}
appearance_drop_precaution <- "DROP TABLE IF EXISTS appearance_type;"
dbExecute(conn, appearance_drop_precaution)
appearance_type = read_csv('appearance_type.csv')
dbWriteTable(conn = conn, name = "appearance_type", 
             value = appearance_type,
             row.names = FALSE, header = TRUE)
check_query_5 <- "SELECT * FROM appearance_type;"
check <- dbGetQuery(conn, check_query_5)
head(check)
```

## Creating Team and Game Tables 

Here are some notes on the normalization choices made with each of these tables:

    team
        The start, end, and sequence columns can be derived from the game level data.
    game
        We have chosen to include all columns for the game log that don't refer to one specific team or player, instead putting those in two appearance tables.
        We have removed the column with the day of the week, as this can be derived from the date.
        We have changed the day_night column to day, with the intention of making this a boolean column. Even though SQLite doesn't support the BOOLEAN type, we can use this when creating our table and SQLite will manage the underlying types behind the scenes (for more on how this works refer to the SQLite documentation. This means that anyone quering the schema of our database in the future understands how that column is intended to be used.

Create the team table with columns, primary key, and foreign key as shown in the schema diagram.

    Select the appropriate type based on the data.
    Insert the data from the team_codes table.
    Write a query to display the first few rows of the table.

```{r}
create_team_command <- "
  CREATE TABLE IF NOT EXISTS team (
      team_id TEXT PRIMARY KEY,
      league_id TEXT,
      city TEXT,
      nickname TEXT,
      franch_id TEXT,
      FOREIGN KEY (league_id) REFERENCES league(league_id)
  );
"
dbExecute(conn, create_team_command)
insert_to_team <- "
  INSERT OR IGNORE INTO team
  SELECT
      team_id,
      league,
      city,
      nickname,
      franch_id
  FROM team_codes;
"
dbExecute(conn, insert_to_team)
check_query_6 <- "SELECT * FROM team LIMIT 5;"
check <- dbGetQuery(conn, check_query_6)
head(check)

```


Create the game table with columns, primary key, and foreign key as shown in the schema diagram.

    Select the appropriate type based on the data.
    Insert the data from the game_log table.
    Write a query to display the first few rows of the table.


```{r}
create_game_command <- "
  CREATE TABLE IF NOT EXISTS game (
      game_id TEXT PRIMARY KEY,
      date TEXT,
      number_of_game INTEGER,
      park_id TEXT,
      length_outs INTEGER,
      day BOOLEAN,
      completion TEXT,
      forefeit TEXT,
      protest TEXT,
      attendance INTEGER,
      legnth_minutes INTEGER,
      additional_info TEXT,
      acquisition_info TEXT,
      FOREIGN KEY (park_id) REFERENCES park(park_id)
  );
"
```

Executing the table 

```{r}
dbExecute(conn, create_game_command)
```

Inserting data from game_log into new table 

```{r}
insert_to_game <- '
  INSERT OR IGNORE INTO game
  SELECT
      game_id,
      date,
      number_of_game,
      park_id,
      length_outs,
      CASE
          WHEN day_night = "D" THEN 1
          WHEN day_night = "N" THEN 0
          ELSE NULL
          END
          AS day,
      completion,
      forefeit,
      protest,
      attendance,
      length_minutes,
      additional_info,
      acquisition_info
  FROM game_log;
'
```

Making the change happen ..

```{r}
dbExecute(conn, insert_to_game)
```

Lets check if the new game table returns everything right 


```{r}
check_query_7 <- "SELECT * FROM game LIMIT 5;"
check <- dbGetQuery(conn, check_query_7)
head(check)
```

Now er will make a table that has optimized number of columns so that we can fit in for example , v error and h error from game_log into one single column called errors. We will call this column 'Team Appearance'. 

As we have two sets of information , home and visitors, we will UNION the two sets so that the rows dont get repeated if they happen to stay in either home or visitor sets. We will create a boolean column called home which will represent if it is a 1 (home) or 0 (visitor) team. As both sets will have new home columns, the UNION clause will ensure that we have one home column in the final table which will consist either 1 or 0. 

```{r}
create_team_appearance_command <- "
  CREATE TABLE IF NOT EXISTS team_appearance (
      team_id TEXT,
      game_id TEXT,
      home BOOLEAN,
      league_id TEXT,
      score INTEGER,
      line_score TEXT,
      at_bats INTEGER,
      hits INTEGER,
      doubles INTEGER,
      triples INTEGER,
      homeruns INTEGER,
      rbi INTEGER,
      sacrifice_hits INTEGER,
      sacrifice_flies INTEGER,
      hit_by_pitch INTEGER,
      walks INTEGER,
      intentional_walks INTEGER,
      strikeouts INTEGER,
      stolen_bases INTEGER,
      caught_stealing INTEGER,
      grounded_into_double INTEGER,
      first_catcher_interference INTEGER,
      left_on_base INTEGER,
      pitchers_used INTEGER,
      individual_earned_runs INTEGER,
      team_earned_runs INTEGER,
      wild_pitches INTEGER,
      balks INTEGER,
      putouts INTEGER,
      assists INTEGER,
      errors INTEGER,
      passed_balls INTEGER,
      double_plays INTEGER,
      triple_plays INTEGER,
      PRIMARY KEY (team_id, game_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id),
      FOREIGN KEY (game_id) REFERENCES game(game_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id)
  );
"
dbExecute(conn, create_team_appearance_command)
insert_to_team_appearance <- "
  INSERT OR IGNORE INTO team_appearance
      SELECT
          h_name,
          game_id,
          1 AS home,
          h_league,
          h_score,
          h_line_score,
          h_at_bats,
          h_hits,
          h_doubles,
          h_triples,
          h_homeruns,
          h_rbi,
          h_sacrifice_hits,
          h_sacrifice_flies,
          h_hit_by_pitch,
          h_walks,
          h_intentional_walks,
          h_strikeouts,
          h_stolen_bases,
          h_caught_stealing,
          h_grounded_into_double,
          h_first_catcher_interference,
          h_left_on_base,
          h_pitchers_used,
          h_individual_earned_runs,
          h_team_earned_runs,
          h_wild_pitches,
          h_balks,
          h_putouts,
          h_assists,
          h_errors,
          h_passed_balls,
          h_double_plays,
          h_triple_plays
      FROM game_log
  
  UNION
  
      SELECT    
          v_name,
          game_id,
          0 AS home,
          v_league,
          v_score,
          v_line_score,
          v_at_bats,
          v_hits,
          v_doubles,
          v_triples,
          v_homeruns,
          v_rbi,
          v_sacrifice_hits,
          v_sacrifice_flies,
          v_hit_by_pitch,
          v_walks,
          v_intentional_walks,
          v_strikeouts,
          v_stolen_bases,
          v_caught_stealing,
          v_grounded_into_double,
          v_first_catcher_interference,
          v_left_on_base,
          v_pitchers_used,
          v_individual_earned_runs,
          v_team_earned_runs,
          v_wild_pitches,
          v_balks,
          v_putouts,
          v_assists,
          v_errors,
          v_passed_balls,
          v_double_plays,
          v_triple_plays
      from game_log;
"
dbExecute(conn, insert_to_team_appearance)
```

Lets check how the table looks like. The minimum of game_id will show the earliest record in the database and max will do the opposite. 


```{r}
check_query_8 <- "
  SELECT * FROM team_appearance
  WHERE game_id = (
                   SELECT MIN(game_id) from game
                  )
     OR game_id = (
                   SELECT MAX(game_id) from game
                  )
  ORDER By game_id, home;
"
check <- dbGetQuery(conn, check_query_8)
head(check)
```

Here, in this check we see that in each game has been recorded two times - record for the home team and the visitor team. 

Now, lets make the person_appearance table . See the schema for the primary and foreign key references. The person_appearance table will be used to store information on appearances in games by managers, players, and umpires as detailed in the appearance_type table.

We'll use a similar technique to insert data as we did with the team_appearance table. A key difference is that we will have to write much larger queries: one for each column instead of one for each team as before. We will need to work out for each column what the appearance_type_id will be by cross-referencing the columns with the appearance_type table.

The primary key for person_appearance will be an integer. Having every column be a compound primary quickly becomes cumbersome when writing queries. In SQLite, if you have an integer primary key and don't specify a value for this column when inserting rows, SQLite will autoincrement this column for you.

```{r}
drop_person_appearance_precaution <- "DROP TABLE IF EXISTS person_appearance"
dbExecute(conn, drop_person_appearance_precaution)
create_person_appearance_command <- "
  CREATE TABLE person_appearance (
      appearance_id INTEGER PRIMARY KEY,
      person_id TEXT,
      team_id TEXT,
      game_id TEXT,
      appearance_type_id,
      FOREIGN KEY (person_id) REFERENCES person(person_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id),
      FOREIGN KEY (game_id) REFERENCES game(game_id),
      FOREIGN KEY (appearance_type_id) REFERENCES appearance_type(appearance_type_id)
  );
"
dbExecute(conn, create_person_appearance_command)
insert_to_person_appearance_2 <- '
  INSERT OR IGNORE INTO person_appearance (
      game_id,
      team_id,
      person_id,
      appearance_type_id
  ) 
      SELECT
          game_id,
          NULL,
          hp_umpire_id,
          "UHP"
      FROM game_log
      WHERE hp_umpire_id IS NOT NULL    
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [1b_umpire_id],
          "U1B"
      FROM game_log
      WHERE "1b_umpire_id" IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [2b_umpire_id],
          "U2B"
      FROM game_log
      WHERE [2b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [3b_umpire_id],
          "U3B"
      FROM game_log
      WHERE [3b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          lf_umpire_id,
          "ULF"
      FROM game_log
      WHERE lf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          rf_umpire_id,
          "URF"
      FROM game_log
      WHERE rf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_manager_id,
          "MM"
      FROM game_log
      WHERE v_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_manager_id,
          "MM"
      FROM game_log
      WHERE h_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_pitcher_id,
          "AWP"
      FROM game_log
      WHERE winning_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score < v_score THEN h_name
              ELSE v_name
              END,
          losing_pitcher_id,
          "ALP"
      FROM game_log
      WHERE losing_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          saving_pitcher_id,
          "ASP"
      FROM game_log
      WHERE saving_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_rbi_batter_id,
          "AWB"
      FROM game_log
      WHERE winning_rbi_batter_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE v_starting_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE h_starting_pitcher_id IS NOT NULL;
'
dbExecute(conn, insert_to_person_appearance_2)
```




When we get to the offensive and defensive positions for both teams, we essentially are performing 36 permutations: 2 (home, away) * 2 (offense + defense) * 9 (9 positions).

Writing out all these permutations will get incredibly tiring, so we'll use a loop along with some regular expression to expedite the process:


For this process we will write a program that will choose the either **h** or **v** and automatically will fill several fields in with these letters so that we know whether they are from home or visitor teams. In the same way, as we have 9 positions in the game, the numbers 1 to 9 will be filled in different fields and we will know where their position was. 

The magic of regular expressions will help this automation process. 



```{r}
for (letter in c("h", "v")) {
  for (num in 1:9) {
    template <- '
      INSERT INTO person_appearance (
          game_id,
          team_id,
          person_id,
          appearance_type_id
      ) 
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "O%f"
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL

      UNION

          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "D" || CAST(%s_player_%f_def_pos AS INT)
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL;
    '
    # replace all of the %s and %f with the correct letter number
    template <- gsub("%s", letter, template, fixed = TRUE)
    template <- gsub("%f", num, template, fixed = TRUE)

    dbExecute(conn, template)
  }
}
```

Lets check person_appearance 


```{r}
check_query_9 <- "
  SELECT * FROM person_appearance
  
"

dbGetQuery(conn, check_query_9)
```

Lets DROP the tables that we createdto hold our un-normalized data 

```{r}
# Check the current status of the db file
dbListTables(conn)
# Iterate through each of the tables
tables <- c("game_log", "park_codes",
            "team_codes", "person_codes")
for (t in tables) {
  drop_command = sprintf("DROP TABLE %s", t)
  dbExecute(conn, drop_command)
}

```

Okay, we ran with some errors in this learning process. This notebook has helped us a lot to understand the ins and outs. We will , however, copy the github solution to a new R notebook and see if we get the desired results. I dont know why game_log isnt there. anyways, lets finish this

```{r}
dbDisconnect(conn)
```

I ran the GIthub solution code in a whole new file but the result looks same...There was a **tinyformat error** after the regular expression chuck was run. It said Too many conversion specifiers in format string. 


Ultimately. the main goal here was to create a clean and optimized database from messy csvs. 
